---
title: "Indonesia Rheumatic Heart Disease Secondary-Prevention Investment Case: Methods, Projected Health Impact and Economic Evaluation (2025–2100)"
date: "`r Sys.Date()`"
author: "University of Washington · World Bank Indonesia RHD Model"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 5,
  fig.width  = 9,
  fig.align  = "center"
)
```

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(DT)
library(scales)
```

```{r paths, message=FALSE, warning=FALSE, echo=FALSE}
# Single source of truth for paths. The project root is resolved relative to
# this .Rmd (analysis/ is one level below the repo root) so the report knits on
# any machine; a hard-coded fall-back is used when the file is run
# chunk-by-chunk outside of knit. This report READS the pipeline's persisted
# output artifacts only -- it never re-runs the model (scripts 06/07/08).
wd <- tryCatch({
  root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
                        winslash = "/", mustWork = TRUE)
  paste0(root, "/")
}, error = function(e)
  "C:/Users/wrgar/OneDrive - UW/02Work/WorldBank-Indonesia/uw-wb-indonesia-rhd/")

wd_data <- paste0(wd, "data/")        # processed inputs + baseline state (from 02-05)
wd_outp <- paste0(wd, "output/")      # model outputs (06) + tables (07/08)
wd_raw  <- paste0(wd, "data-raw/")    # GBD 2023 epi rates (for historical validation)
```

```{r helpers, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Number / currency formatters -------------------------------------------
fmt_n   <- function(x, d = 0) formatC(round(x, d), format = "f", digits = d, big.mark = ",")
fmt_m   <- function(x, d = 2) formatC(x / 1e6, format = "f", digits = d, big.mark = ",")
fmt_k   <- function(x, d = 0) formatC(x / 1e3, format = "f", digits = d, big.mark = ",")
fmt_pct <- function(x, d = 1) paste0(formatC(100 * x, format = "f", digits = d), "%")
fmt_usd <- function(x, d = 0) paste0("$", formatC(round(x, d), format = "f", digits = d, big.mark = ","))
usd_b   <- function(x, d = 2) paste0("$", formatC(x / 1e9, format = "f", digits = d, big.mark = ","), " billion")
usd_m   <- function(x, d = 1) paste0("$", formatC(x / 1e6, format = "f", digits = d, big.mark = ","), " million")

# ---- Scenario metadata (authoritative 'scenario' column from 06) ------------
# 06 writes exactly two arms per location, distinguished by 'scenario':
#   ref = reference (care cascade held at its baseline coverage);
#   sap = scale-up of the screening -> diagnosis -> optimal-treatment cascade.
scenario_levels <- c("ref", "sap")
scenario_labels <- c(
  ref = "Reference (cascade held at baseline)",
  sap = "SAP scale-up (screening + diagnosis + prophylaxis)"
)
scenario_labels_short <- c(ref = "Reference", sap = "SAP scale-up")
scenario_cols <- c(ref = "#4d4d4d", sap = "#2166ac")
reduction_col <- "#b2182b"
sex_cols      <- c(Female = "#D81B60", Male = "#1E88E5")

# ---- GBD single-year-age -> GBD-group label (mirrors 03/04 age_to_gbd_group) -
# Used to fold the model's single-year ages into the 22 GBD age groups so
# model and GBD age profiles share one comparable age axis.
age_to_grp <- function(a) {
  fcase(
    a < 1,  "<1 year",
    a < 2,  "12-23 months",
    a < 5,  "2-4 years",
    a < 10, "5-9 years",
    a < 15, "10-14 years", a < 20, "15-19 years", a < 25, "20-24 years",
    a < 30, "25-29 years", a < 35, "30-34 years", a < 40, "35-39 years",
    a < 45, "40-44 years", a < 50, "45-49 years", a < 55, "50-54 years",
    a < 60, "55-59 years", a < 65, "60-64 years", a < 70, "65-69 years",
    a < 75, "70-74 years", a < 80, "75-79 years", a < 85, "80-84 years",
    a < 90, "85-89 years", a < 95, "90-94 years",
    default = "95+ years"
  )
}
# Ordered group levels + a numeric midpoint for each (age-profile x-axis).
grp_levels <- c("<1 year", "12-23 months", "2-4 years", "5-9 years",
                "10-14 years", "15-19 years", "20-24 years", "25-29 years",
                "30-34 years", "35-39 years", "40-44 years", "45-49 years",
                "50-54 years", "55-59 years", "60-64 years", "65-69 years",
                "70-74 years", "75-79 years", "80-84 years", "85-89 years",
                "90-94 years", "95+ years")
grp_mid <- setNames(
  c(0.5, 1.5, 3, 7, 12, 17, 22, 27, 32, 37, 42, 47, 52, 57, 62, 67, 72, 77,
    82, 87, 92, 97),
  grp_levels)

# Common ggplot theme tweaks used throughout.
base_theme <- function(sz = 13)
  theme_minimal(base_size = sz) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
```

```{r load-model-output, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Load per-location model outputs written by 06_run_prevention_model.R ---
# One file per location: output/out_model/<location>.rds, each a list with
# $wsd (aggregate well-sick-dead view, sick = A + B + C + D), $stages (the four
# A/B/C/D living stocks + every transition flow + the surgery-service trace +
# cascade coverages), $diag and $meta. 07 compiles these into the tables in
# output/tables/.
out_model_path <- paste0(wd_outp, "out_model/")
mfiles <- list.files(out_model_path, pattern = "\\.rds$", full.names = TRUE)
if (length(mfiles) == 0)
  stop(paste0("No model outputs found in:\n  ", out_model_path,
              "\nRun scripts/06_run_prevention_model.R first."))

wsd <- rbindlist(lapply(mfiles, function(f) as.data.table(readRDS(f)$wsd)),
                 use.names = TRUE, fill = TRUE)
stg <- rbindlist(lapply(mfiles, function(f) as.data.table(readRDS(f)$stages)),
                 use.names = TRUE, fill = TRUE)
meta  <- readRDS(mfiles[1])$meta
diag1 <- readRDS(mfiles[1])$diag

# Keep recognised scenarios + impose ordering.
wsd <- wsd[scenario %in% scenario_levels]
stg <- stg[scenario %in% scenario_levels]
wsd[, scenario := factor(scenario, levels = scenario_levels)]
stg[, scenario := factor(scenario, levels = scenario_levels)]

# ---- Key dimensions ---------------------------------------------------------
LOCS        <- sort(unique(as.character(wsd$location)))
n_loc       <- length(LOCS)
year_min    <- min(wsd$year); year_max <- max(wsd$year)
base_year   <- meta$base_year                       # 2025 seed / jump-off year
ramp_start  <- meta$ramp_start; ramp_end <- meta$ramp_end
proj_years  <- paste0(base_year, "–", year_max)
sim_horizon <- paste0(base_year + 1L, "–", year_max)   # 2026-2100 (task horizon)
CAUSE       <- unique(as.character(wsd$cause))[1]
scenarios_present <- levels(droplevels(wsd$scenario))
```

```{r load-baseline-state, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Baseline state assembled by 05_build_baseline.R ------------------------
# Holds, per location, the A/B/C/D model INPUTS that 06 consumes:
#   $transitions  annual A/B/C/D transition probabilities
#   $p_rhd_death  per-stage RHD-specific death probabilities c(A,B,C,D)
#   $effects      sap_rrr_rhd_death, eff_surgery_C_to_D, eff_surgery_D_to_rhd_death
#   $surgery      frac_C/D_requiring_surgery + per-arm surgery coverage
#   $coverage     REALISED cascade + surgery coverage trajectories per scenario
#                 ($ref/$sap each: screen, diagnosis, treatment, eff_diagnosis,
#                  eff_treatment, surgery), plus the optional screening-age mask.
bs_path <- paste0(wd_data, "baseline_state.rds")
have_bs <- file.exists(bs_path)
if (have_bs) {
  bstate  <- readRDS(bs_path)
  loc1    <- bstate$locations[1]
  tr      <- bstate$states[[loc1]]$transitions   # 7 A/B/C/D transition probabilities
  pdth    <- bstate$states[[loc1]]$p_rhd_death    # per-stage RHD death probs c(A,B,C,D)
  eff     <- bstate$states[[loc1]]$effects        # sap_rrr_rhd_death + surgery RRRs
  srg     <- bstate$states[[loc1]]$surgery        # surgery-service fractions + coverage
  cov     <- bstate$states[[loc1]]$coverage       # ref/sap cascade + surgery trajectories
  cov_years <- bstate$meta$years
  stage_split <- bstate$meta$stage_split          # A/B/C/D prevalent-pool seed split
} else {
  bstate <- tr <- pdth <- eff <- srg <- cov <- NULL
  loc1 <- LOCS[1]; cov_years <- NULL; stage_split <- meta$stage_split
}
```

```{r load-stage-flow-tables, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Compiled stage + flow long tables from 07_make_outputs.R ---------------
# Preferred source for stage-specific quantities (07 already provides the exact
# per-stage prevalence and the annual flow / surgery-trace volumes). Optional:
# the report degrades to deriving these from $stages ('stg') if absent.
stage_path <- paste0(wd_outp, "tables/rhd_stage_model_long.rds")
flow_path  <- paste0(wd_outp, "tables/rhd_annual_flows.rds")
have_stage_tab <- file.exists(stage_path)
have_flow_tab  <- file.exists(flow_path)
STAGES <- c("A", "B", "C", "D")
stage_labels <- c(A = "A · minimal / early", B = "B · mild established",
                  C = "C · advanced, no complications", D = "D · advanced with complications")
stage_cols   <- c(A = "#8CBFDA", B = "#4E97C7", C = "#E4A93C", D = "#B2182B")

if (have_stage_tab) {
  stage_long <- as.data.table(readRDS(stage_path))
} else {
  # derive from $stages: melt the four stocks and attach prevalence per 1,000.
  stage_long <- melt(stg[, .(scenario, age, sex, year, location, pop, A, B, C, D)],
                     id.vars = c("scenario", "age", "sex", "year", "location", "pop"),
                     measure.vars = STAGES, variable.name = "stage", value.name = "cases")
  stage_long[, `:=`(prevalence = fifelse(pop > 0, cases / pop, 0),
                    prevalence_per_1000 = fifelse(pop > 0, cases / pop * 1000, 0))]
  stage_long[, pop := NULL]
}
stage_long[, scenario := factor(as.character(scenario), levels = scenario_levels)]
stage_long[, stage := factor(as.character(stage), levels = STAGES)]

flows <- if (have_flow_tab) as.data.table(readRDS(flow_path)) else copy(stg)
flows[, scenario := factor(as.character(scenario), levels = scenario_levels)]
```

```{r load-econ, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Economic evaluation bundle from 08_economic_evaluation.R ---------------
# rhd_economic_results.rds = list($summary, $budget_impact, $incremental,
# $params, $discount_base_year). All monetary values live in 08 only.
er_path <- paste0(wd_outp, "tables/rhd_economic_results.rds")
have_econ <- file.exists(er_path)
if (have_econ) {
  er        <- readRDS(er_path)
  econ_sum  <- as.data.table(er$summary)
  budget    <- as.data.table(er$budget_impact)
  incr      <- as.data.table(er$incremental)
  eparams   <- er$params
  disc_base <- er$discount_base_year
} else {
  er <- econ_sum <- budget <- incr <- eparams <- disc_base <- NULL
}
```

```{r load-gbd, message=FALSE, warning=FALSE, echo=FALSE}
# ---- GBD 2023 epidemiology (1990-2023) for the historical validation panels -
# From data-raw/temp_baseline_rates_gbd.rds (01_prepare_inputs.R): Indonesia,
# causes {All causes, Rheumatic heart disease}, measures {Deaths, Prevalence,
# Incidence}, metrics {Number, Rate}, both sexes. data-raw/ is not tracked by
# git, so this is treated as an OPTIONAL artifact -- the report degrades
# gracefully to model-only series if it is absent.
gbd_try <- c(paste0(wd_raw,  "temp_baseline_rates_gbd.rds"),
             paste0(wd_data, "temp_baseline_rates_gbd.rds"))
gbd_path <- gbd_try[file.exists(gbd_try)][1]
have_gbd <- !is.na(gbd_path)
if (have_gbd) {
  gbd <- as.data.table(readRDS(gbd_path))
  gbd <- gbd[location_name == loc1]
}

# Observed single-year population 1990-2024 (02_build_demography.R) as the
# denominator for GBD crude rates.
po_path <- paste0(wd_data, "pop_observed_1990_2024.rds")
have_pop <- file.exists(po_path)
if (have_pop) {
  pop_obs <- as.data.table(readRDS(po_path))
  pop_obs <- pop_obs[location == loc1]
  gbd_pop_yr <- pop_obs[, .(pop = sum(Nx)), by = year]   # total pop by year
} else {
  gbd_pop_yr <- NULL
}

# Helper: GBD annual count (Number metric, summed over age & sex) for a
# (cause, measure) pair -> data.table(year, count).
gbd_count_yr <- function(cause, measure) {
  if (!have_gbd) return(NULL)
  gbd[cause_name == cause & measure_name == measure & metric_name == "Number",
      .(count = sum(val)), by = year][order(year)]
}
```

```{r headline-precompute, message=FALSE, warning=FALSE, echo=FALSE}
# Cumulative RHD deaths, by scenario, over the full horizon (from $wsd `dead`).
scn_deaths <- wsd[, .(dead = sum(dead)), by = scenario]
deaths_ref <- scn_deaths[scenario == "ref", dead]
deaths_sap <- scn_deaths[scenario == "sap", dead]
deaths_averted_cum <- deaths_ref - deaths_sap

# Economic headlines (guarded).
if (have_econ) {
  tot <- econ_sum[location == "TOTAL"]
  if (!nrow(tot)) tot <- econ_sum[.N]     # single-location fall-back
  bcr_hl        <- tot$bcr
  netben_hl     <- tot$net_benefit
  inccost_hl    <- tot$total_cost
  benefit_hl    <- tot$total_benefit
  cpd_hl        <- tot$cost_per_death
  cpdaly_hl     <- tot$cost_per_daly
  deaths_av_hl  <- tot$deaths_averted
} else {
  deaths_av_hl <- deaths_averted_cum
}
```

---

# Background

This report documents the **Indonesia rheumatic heart disease (RHD)
secondary-prevention investment-case model** implemented in the numbered
pipeline `scripts/06_run_prevention_model.R` (the runner),
`scripts/07_make_outputs.R` (the standard long table) and
`scripts/08_economic_evaluation.R` (the economics), and presents its baseline,
projected and economic outputs. It reads **only the pipeline's persisted output
artifacts** — the per-location `.rds` files in `output/out_model/` and the
tables in `output/tables/` — together with the GBD 2023 epidemiology used for
historical validation. It does **not** re-run the model.

RHD is an acquired valvular heart disease that begins with acute rheumatic
fever following group A streptococcal infection in childhood and progresses,
over years, from a minimal (subclinical) valve lesion through advancing valve
disease to heart failure with complications, and in advanced disease to valve
surgery or death. It remains a major, largely preventable cause of
cardiovascular death among young people in Indonesia. The
**secondary-prevention investment case** evaluated here couples
**echocardiographic screening** and **diagnosis** (to detect and confirm RHD)
with **secondary antibiotic prophylaxis (SAP) / optimal treatment** — long-acting
penicillin that suppresses recurrent streptococcal infection and thereby reduces
RHD-specific mortality across the disease course.

The central question is:

> *How many RHD deaths does scaling up echo screening and secondary antibiotic
> prophylaxis avert in Indonesia over the century, and does the monetised
> health benefit justify the incremental cost?*

The engine is a **discrete-time, matrix-form (age × sex) Markov cohort model**
whose living RHD population is resolved into the four **World Heart Federation
stages** — **A** (minimal / early), **B** (mild established), **C** (advanced
without complications) and **D** (advanced with complications, heart failure) —
run once per **scenario**:

- a **reference** (`ref`) arm, with the care cascade held at its baseline
  coverage, and
- a **SAP scale-up** (`sap`) arm, in which screening, diagnosis and optimal
  treatment rise to their targets.

The model is structured after **Coates et al., *Lancet Global Health* 2021**
(the global RHD investment-case framework), adapted to Indonesian demography and
GBD 2023 epidemiology and to the WHF A/B/C/D staging of RHD.

```{r bg-inline, echo=FALSE, results='asis'}
cat(sprintf(
"Over the projection horizon (%s), across %d location(s), scaling up secondary
antibiotic prophylaxis is projected to avert approximately **%s RHD deaths**
relative to the reference arm.",
  proj_years, n_loc, fmt_n(deaths_averted_cum)))
if (have_econ) cat(sprintf(
" The scale-up carries a discounted incremental cost of **%s**, a monetised
benefit of **%s**, a **net benefit of %s**, and a **benefit–cost ratio of
%.2f** (about **%s per DALY averted**). Full economics are in the
[economic evaluation](#economic-evaluation).",
  usd_b(inccost_hl), usd_b(benefit_hl), usd_b(netben_hl), bcr_hl, fmt_usd(cpdaly_hl)))
```

> **A note on terminology.** Because SAP reduces **RHD-specific mortality** in
> every living stage (it acts on case fatality, not on incidence), it genuinely
> removes RHD deaths from the horizon rather than merely delaying them. We
> therefore report **deaths averted** (reference deaths − scale-up deaths). One
> consequence, made explicit in the [results](#results), is that averting
> deaths leaves more people **alive with RHD (predominantly in the earlier
> stages)**, so modelled *prevalent* cases can be slightly **higher** under the
> scale-up even as deaths fall.

---

# Model inputs

The parameters below are consumed *upstream* of the runner by
`05_build_baseline.R` (natural-history, effect, surgery-service and cascade
inputs) and by `08_economic_evaluation.R` (economic inputs). **Every value in the
table is read programmatically from the artifacts** — the natural-history,
effect, surgery and coverage values from `data/baseline_state.rds`
(`$states[[loc]]$transitions`, `$p_rhd_death`, `$effects`, `$surgery`,
`$coverage`) and the economic values from
`output/tables/rhd_economic_results.rds$params` — so the report can never drift
from the model. Provenance tags follow the pipeline's own convention:
**`[LIT]`** literature value, **`[CALIBRATE]`** tuned to setting/appendix,
**`[PAPER]`** article main-text / user control.

```{r tbl-inputs, message=FALSE, warning=FALSE, echo=FALSE}
inp <- data.table(Group = character(), Parameter = character(),
                  Value = character(), Provenance = character())

add_row <- function(g, p, v, prov)
  inp <<- rbind(inp, data.table(Group = g, Parameter = p, Value = v, Provenance = prov))

if (have_bs) {
  # A/B/C/D annual transition probabilities
  g <- "A/B/C/D transition probabilities (per year)"
  add_row(g, "A → No RHD (regression)", fmt_pct(tr$p_A_to_no_rhd, 1), "[CALIBRATE]")
  add_row(g, "A → B (progression)",     fmt_pct(tr$p_A_to_B, 1),      "[CALIBRATE]")
  add_row(g, "B → A (regression)",      fmt_pct(tr$p_B_to_A, 1),      "[CALIBRATE]")
  add_row(g, "B → C (progression)",     fmt_pct(tr$p_B_to_C, 1),      "[CALIBRATE]")
  add_row(g, "C → B (regression)",      fmt_pct(tr$p_C_to_B, 1),      "[CALIBRATE]")
  add_row(g, "C → D (progression)",     fmt_pct(tr$p_C_to_D, 1),      "[CALIBRATE]")
  add_row(g, "D → C (regression)",      fmt_pct(tr$p_D_to_C, 1),      "[CALIBRATE]")

  # Per-stage RHD-specific mortality (untreated)
  g <- "Per-stage RHD-specific mortality (per year, untreated)"
  add_row(g, "Stage A ($p^{d}_{A}$)", fmt_pct(pdth[["A"]], 2), "[CALIBRATE]")
  add_row(g, "Stage B ($p^{d}_{B}$)", fmt_pct(pdth[["B"]], 2), "[CALIBRATE]")
  add_row(g, "Stage C ($p^{d}_{C}$)", fmt_pct(pdth[["C"]], 2), "[CALIBRATE]")
  add_row(g, "Stage D ($p^{d}_{D}$)", fmt_pct(pdth[["D"]], 2), "[CALIBRATE]")

  # Intervention effects (relative risk reductions)
  g <- "Intervention effects (relative risk reduction)"
  add_row(g, "SAP on RHD-specific mortality, all stages ($\\rho_{sap}$)", fmt_pct(eff$sap_rrr_rhd_death, 0), "[PAPER]")
  add_row(g, "Surgery on C → D progression ($\\rho^{surg}_{CD}$)",        fmt_pct(eff$eff_surgery_C_to_D, 0), "[LIT]")
  add_row(g, "Surgery on D → RHD death ($\\rho^{surg}_{Dd}$)",            fmt_pct(eff$eff_surgery_D_to_rhd_death, 0), "[LIT]")

  # Surgery service (held equal in both arms)
  g <- "Surgery service (both arms)"
  add_row(g, "Fraction of stage C requiring surgery / year", fmt_pct(srg$frac_C_requiring_surgery, 1), "[CALIBRATE]")
  add_row(g, "Fraction of stage D requiring surgery / year", fmt_pct(srg$frac_D_requiring_surgery, 1), "[CALIBRATE]")
  add_row(g, "Surgery coverage (both arms)", fmt_pct(cov$ref$surgery[1], 0), "[CALIBRATE]")

  # Care cascade (cumulative population coverage; scale-up ramps, reference flat)
  g <- "Care cascade (cumulative coverage)"
  add_row(g, "Screening — baseline → 2050 target",
          sprintf("%s → %s", fmt_pct(cov$ref$screen[1], 0),    fmt_pct(max(cov$sap$screen), 0)),    "[PAPER]")
  add_row(g, "Diagnosis — baseline → 2050 target",
          sprintf("%s → %s", fmt_pct(cov$ref$diagnosis[1], 0), fmt_pct(max(cov$sap$diagnosis), 0)), "[PAPER]")
  add_row(g, "Optimal treatment (SAP) — baseline → 2050 target",
          sprintf("%s → %s", fmt_pct(cov$ref$treatment[1], 0), fmt_pct(max(cov$sap$treatment), 0)), "[PAPER]")
  add_row(g, "Scale-up ramp window", sprintf("%d → %d", ramp_start, ramp_end), "[PAPER]")
  add_row(g, "Incidence secular trend (per year)", sprintf("%.3f", meta$incidence_trend), "[CALIBRATE]")

  # Prevalent-pool stage split at the seed year
  g <- "Prevalent-pool stage split (seed year)"
  add_row(g, "Stage A share", fmt_pct(stage_split[["A"]], 1), "[LIT]")
  add_row(g, "Stage B share", fmt_pct(stage_split[["B"]], 1), "[LIT]")
  add_row(g, "Stage C share", fmt_pct(stage_split[["C"]], 1), "[LIT]")
  add_row(g, "Stage D share (complicated-RHD fraction)", fmt_pct(stage_split[["D"]], 1), "[CALIBRATE]")
}

if (have_econ) {
  g <- "Economic parameters (08; base-year US$)"
  add_row(g, "Discount rate (costs & benefits)", fmt_pct(eparams$disc_rate, 0), "[PAPER]")
  add_row(g, sprintf("GDP per capita (%d)", eparams$gdp_pc_base_year), fmt_usd(eparams$gdp_pc_base), "[LIT]")
  add_row(g, "Real per-capita GDP growth", fmt_pct(eparams$gdp_growth, 0), "[CALIBRATE]")
  add_row(g, "Value of a statistical life (× GDP-pc)", sprintf("%g×", eparams$vsl_mult), "[PAPER]")
  add_row(g, "Undiscounted DALYs per RHD death", sprintf("%g", eparams$dalys_per_death), "[CALIBRATE]")
  add_row(g, "Echo screening cost / person screened", fmt_usd(eparams$cost_screen_per_person), "[PAPER]")
  add_row(g, "Optimal-treatment (SAP) cost / person-year", fmt_usd(eparams$cost_sap_per_year), "[PAPER]")
  add_row(g, "Valve surgery + first-year post-op cost / surgery", fmt_usd(eparams$cost_surgery), "[LIT]")
  add_row(g, "Discount base year", sprintf("%d", disc_base), "[PAPER]")
}

if (nrow(inp)) {
  rl  <- rle(inp$Group)
  idx <- setNames(rl$lengths, rl$values)
  disp <- inp[, .(Parameter, Value, Provenance)]
  kbl <- kable(disp,
        caption = "Table 1. Key model input parameters, by provenance",
        align = c("l", "r", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,   # <-- set to FALSE to respect explicit widths
                font_size = 13) |>
  column_spec(1, width = "7cm") |>    # <-- adjust the value as needed
  row_spec(0, bold = TRUE)

# then your pack_rows loop
for (i in seq_along(idx))
  kbl <- pack_rows(kbl, names(idx)[i],
                   start_row = sum(idx[seq_len(i - 1)]) + 1L,
                   end_row   = sum(idx[seq_len(i)]))
kbl
} else {
  cat("Input artifacts unavailable (baseline_state.rds / rhd_economic_results.rds not found).")
}
```

```{r fig-coverage, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=4}
# Realised care-cascade coverage: the scale-up ramps screening -> diagnosis ->
# optimal treatment to their 2050 targets, while the reference arm holds every
# metric flat at its baseline. Surgery coverage is held equal in both arms.
if (have_bs) {
  metric_lab <- c(screen = "Screening", diagnosis = "Diagnosis",
                  treatment = "Optimal treatment (SAP)")
  metric_col <- c(screen = "#4E79A7", diagnosis = "#59A14F", treatment = "#E15759")
  covdt <- rbindlist(lapply(names(metric_lab), function(m)
    rbindlist(list(
      data.table(year = cov_years, coverage = cov$ref[[m]], metric = m, scenario = "ref"),
      data.table(year = cov_years, coverage = cov$sap[[m]], metric = m, scenario = "sap")))))
  covdt[, `:=`(metric = factor(metric, levels = names(metric_lab)),
               scenario = factor(scenario, levels = scenario_levels))]
  ggplot(covdt, aes(year, coverage, color = metric, linetype = scenario)) +
    annotate("rect", xmin = ramp_start, xmax = ramp_end, ymin = -Inf, ymax = Inf,
             alpha = 0.08, fill = "#2166ac") +
    geom_line(linewidth = 1.1) +
    scale_color_manual(values = metric_col, labels = metric_lab, name = NULL) +
    scale_linetype_manual(values = c(ref = "dashed", sap = "solid"),
                          labels = scenario_labels_short, name = NULL) +
    scale_y_continuous(labels = percent, limits = c(0, NA)) +
    labs(title = "Realised care-cascade coverage",
         subtitle = paste0("Solid = scale-up, dashed = reference (held flat); shaded band = ",
                           "ramp window (", ramp_start, "–", ramp_end, ")"),
         x = "Year", y = "Cumulative coverage") +
    base_theme() + theme(legend.position = "bottom")
}
```

---

# Methods

## Overall structure

```{r fig-model-schematic, echo=FALSE, out.width="100%", fig.align="center", fig.cap="Figure M1. Rheumatic heart disease A/B/C/D model structure. An age–sex structured annual cohort for Indonesia. Incident RHD enters at stage A; the four living stages progress and regress along A ↔ B ↔ C with forward progression C → D; stage D feeds RHD death; competing background (non-RHD) mortality acts on every living stage. Secondary antibiotic prophylaxis (SAP) reduces stage-specific RHD mortality; surgery is a clinical service that reduces C → D progression and D → RHD death (it does not remove patients from a stage). Survivors age one year per cycle, with an open terminal 100+ group. Acute rheumatic fever is not represented as a health state."}
schematic <- file.path(wd, "docs", "rhd_abcd_state_transition.png")
if (file.exists(schematic)) knitr::include_graphics(schematic)
```

Let $t$ index calendar years over the horizon (`r proj_years`) and let each
quantity be an **age × sex matrix** (single-year ages $0,\dots,100$ with an open
terminal $100+$ group; sexes {Female, Male}). For the single location the model
tracks a susceptible **No RHD** stock, four living RHD stocks — **A** (minimal /
early), **B** (mild established), **C** (advanced without complications) and
**D** (advanced with complications) — and two absorbing outcomes, **RHD death**
and **other-cause death**. Markov cycles are implemented as elementwise
($\odot$) matrix operations, one per year.

Three data-fed rate arrays drive the dynamics: the **calibrated incidence
probability** $IR_t$ (No RHD → A; from `04`, with a secular trend baked in), the
**background (non-RHD) competing mortality** $\mu$ (held fixed at its base-year
pattern, the same age–sex risk for every living stage), and the population $P_t$
(WPP2024 medium variant from `02`). RHD-specific mortality is carried by the
**per-stage death probabilities** $p^{d}_{A},\dots,p^{d}_{D}$ (Table 1); the
calibrated case-fatality from `04` is used only as a base-year anchor.

**Initialisation (seed year `r base_year`).** The prevalent RHD pool is seeded
from GBD 2023 prevalence and split across the four stages by the tagged
`stage_split` vector (A/B/C from a multi-state severity study, normalised to the
non-D share; D from the complicated-RHD fraction), so that
$A_{0}+B_{0}+C_{0}+D_{0} = \text{PREV}\odot P_{0}$.

**Annual recurrence.** Writing $A_t, B_t, C_t, D_t$ for the stage stocks entering
year $t$, $c^{tx}_t$ for the effective optimal-treatment coverage and
$c^{surg}_t$ for surgery coverage, `06_run_prevention_model.R` first forms the
incident inflow, the SAP mortality multiplier, and the surgery reach:

$$
\begin{aligned}
S_t &= A_t + B_t + C_t + D_t, \qquad W_t = \max\!\left(P_t - S_t,\; 0\right) \\[4pt]
N_t &= W_t \odot IR_t \qquad (\text{new incident RHD into A; } \text{eff\_ir}=1)\\[4pt]
\text{eff\_cf}_t &= 1 - \rho_{\text{sap}}\, c^{tx}_t \qquad(\text{SAP: RHD-mortality multiplier, all stages})\\[4pt]
r^{C}_t &= f_{C}\, c^{surg}_t, \qquad r^{D}_t = f_{D}\, c^{surg}_t \qquad(\text{population-average surgery reach})
\end{aligned}
$$

Surgery (a **service**, not a state) enters only through two risk multipliers —
on $C\!\to\!D$ progression and on $D\!\to\!$ RHD death — combined
multiplicatively with the SAP mortality effect on stage D:

$$
\tilde p_{CD} = p_{CD}\big(1 - \rho^{surg}_{CD}\, r^{C}_t\big), \qquad
\tilde p^{d}_{D} = p^{d}_{D}\,\text{eff\_cf}_t\,\big(1 - \rho^{surg}_{Dd}\, r^{D}_t\big).
$$

The stage stocks then advance, each "stay" term floored at zero and every stage
subject to competing background mortality $\mu$ (regression terms $p_{BA},p_{CB},
p_{DC}$ and the No-RHD exit $p_{A0}$ included):

$$
\begin{aligned}
A^{+}_t &= \max\!\big(A_t(1 - p_{A0} - p_{AB} - p^{d}_{A}\text{eff\_cf}_t - \mu),\,0\big) + B_t\,p_{BA} + N_t \\[4pt]
B^{+}_t &= \max\!\big(B_t(1 - p_{BA} - p_{BC} - p^{d}_{B}\text{eff\_cf}_t - \mu),\,0\big) + A_t\,p_{AB} + C_t\,p_{CB} \\[4pt]
C^{+}_t &= \max\!\big(C_t(1 - p_{CB} - \tilde p_{CD} - p^{d}_{C}\text{eff\_cf}_t - \mu),\,0\big) + B_t\,p_{BC} + D_t\,p_{DC} \\[4pt]
D^{+}_t &= \max\!\big(D_t(1 - p_{DC} - \tilde p^{d}_{D} - \mu),\,0\big) + C_t\,\tilde p_{CD}
\end{aligned}
$$

The surgery trace is recorded on the start-of-cycle stocks,
$\text{surg}^{C}_t = C_t f_{C} c^{surg}_t$ and $\text{surg}^{D}_t = D_t f_{D} c^{surg}_t$
(with $\text{surg}^{C}_t \le C_t f_{C}$, and likewise for D): it feeds the cost
model and the risk multipliers above but is **never subtracted from any stage
stock**. Year-end RHD (cause-specific) and all-cause deaths are

$$
\begin{aligned}
D^{\text{rhd}}_t &= \text{eff\_cf}_t\big(A_t\,p^{d}_{A} + B_t\,p^{d}_{B} + C_t\,p^{d}_{C}\big) + D_t\,\tilde p^{d}_{D} \\[4pt]
A^{\text{mx}}_t &= D^{\text{rhd}}_t \;+\; \mu \odot P_t .
\end{aligned}
$$

Finally the surviving stocks are **aged one year** ($A_{t+1} = \text{shift}(A^{+}_t)$,
and likewise for $B, C, D$), with the terminal $100+$ group absorbing survivors
who age past it. Every quantity is floored at zero. The aggregate reported "sick"
stock is $S = A + B + C + D$ by construction, so the stage-and-flow table and the
aggregate view reconcile exactly.

## What the interventions change

Only the **care cascade differs between the two arms**; surgery coverage and its
effects are held equal in both, so the incremental impact isolates *secondary
prevention*. The reported effect multipliers are, on the $[0,1]$ scale (1 = no
reduction):

$$
\boxed{\;\text{eff\_ir} = 1\;}\qquad\qquad
\boxed{\;\text{eff\_cf}_t = 1 - \rho_{\text{sap}}\, c^{tx}_t\;}
$$

- **`eff_ir` = 1 everywhere** — secondary prevention does **not** reduce RHD
  *incidence*; new cases $N_t = W_t \odot IR_t$ enter stage A governed by the
  calibrated incidence array alone. (A negligible incidence difference between
  arms arises only because the No-RHD pool $W_t = P_t - S_t$ shifts slightly as
  survival improves.)
- **`eff_cf`** — SAP scales the RHD-specific death probability of **every living
  stage** by $1 - \rho_{\text{sap}} c^{tx}_t$, where $c^{tx}_t$ is the *effective
  optimal-treatment coverage* (the cascade minimum of screening, diagnosis and
  treatment). This is exactly the multiplier the engine applies, so the reported
  `eff_cf` column is faithful to the mechanics. In the reference arm
  ($c^{tx}=$ `r if (have_bs) fmt_pct(cov$ref$eff_treatment[1],0) else "4%"`)
  `eff_cf` $\approx$ `r sprintf("%.4f", 1 - (if (have_bs) eff$sap_rrr_rhd_death else .55) * (if (have_bs) cov$ref$eff_treatment[1] else .04))`;
  under the scale-up it falls to
  `r sprintf("%.4f", 1 - (if (have_bs) eff$sap_rrr_rhd_death else .55) * (if (have_bs) max(cov$sap$eff_treatment) else .65))`
  at target coverage.
- **Surgery** additionally reduces $C\!\to\!D$ progression and $D\!\to\!$ RHD
  death in proportion to its population-average reach ($f\cdot c^{surg}$).
  Because that reach is identical in both arms, surgery contributes to the level
  of disease and death in each arm but **not** to the *incremental* effect.

## Causal chain

```{r flow-diagram, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=3.1}
# No RHD -> A <-> B <-> C -> D -> RHD death, with the intervention levers labelled.
boxes <- data.table(
  x = 1:6, y = 1,
  fill = c("#eef2f7", "#8CBFDA", "#4E97C7", "#E4A93C", "#B2182B", "#5b5b5b"),
  label = c("No RHD", "A\nminimal", "B\nmild", "C\nadvanced", "D\ncomplicated",
            "RHD\ndeath"))
fwd <- data.table(x = (1:5) + 0.47, xend = (2:6) - 0.47, y = 1.03, yend = 1.03,
  lab = c("incident RHD\n(IR)", "progression", "progression",
          "progression\nsurgery ↓ C→D", "RHD mortality\nSAP ↓ · surgery ↓ (D)"))
reg <- data.table(x = (3:4) + 0.30, xend = (2:3) + 0.70, y = 0.86, yend = 0.86,
  lab = c("regression", "regression"))
ggplot() +
  geom_tile(data = boxes, aes(x = x, y = y, fill = fill), width = 0.92, height = 0.58,
            color = "grey30", linewidth = 0.6, alpha = 0.85) +
  scale_fill_identity() +
  geom_text(data = boxes, aes(x = x, y = y, label = label), size = 3.0, lineheight = 0.95) +
  # forward progression + incidence + death
  geom_segment(data = fwd, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.16, "cm"), type = "closed"),
               linewidth = 0.5, color = "grey30") +
  geom_text(data = fwd, aes(x = (x + xend) / 2, y = 1.26, label = lab),
            size = 2.35, lineheight = 0.9, color = "grey30") +
  # regression A <- B, B <- C (bidirectional A<->B<->C)
  geom_segment(data = reg, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.13, "cm"), type = "closed"),
               linewidth = 0.4, color = "#6a51a3", linetype = "22") +
  geom_text(data = reg, aes(x = (x + xend) / 2, y = 0.74, label = lab),
            size = 2.2, color = "#6a51a3") +
  coord_cartesian(xlim = c(0.5, 6.5), ylim = c(0.68, 1.42)) +
  labs(title = paste0("No RHD → A ↔ B ↔ C → D → RHD death",
                      "  (SAP reduces stage mortality; surgery is a service)")) +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 11))
```

## Scenarios

```{r tbl-scenarios, message=FALSE, warning=FALSE, echo=FALSE}
scn_tbl <- data.table(
  Scenario = scenario_labels[scenario_levels],
  `Care cascade (screening / diagnosis / optimal treatment)` = c(
    if (have_bs) sprintf("held at baseline (%s / %s / %s)",
                         fmt_pct(cov$ref$screen[1], 0), fmt_pct(cov$ref$diagnosis[1], 0),
                         fmt_pct(cov$ref$treatment[1], 0)) else "held at baseline",
    if (have_bs) sprintf("%s / %s / %s → %s / %s / %s (by %d)",
                         fmt_pct(cov$sap$screen[1], 0), fmt_pct(cov$sap$diagnosis[1], 0),
                         fmt_pct(cov$sap$treatment[1], 0),
                         fmt_pct(max(cov$sap$screen), 0), fmt_pct(max(cov$sap$diagnosis), 0),
                         fmt_pct(max(cov$sap$treatment), 0), ramp_end) else "scaled up to targets"),
  `Surgery coverage` = "held equal in both arms",
  `In outputs` = fifelse(scenario_levels %in% scenarios_present, "yes", "no"))
kable(scn_tbl,
      caption = "Table 2. The two modelled arms. Only the care cascade differs; surgery is held equal.",
      align = c("l", "l", "l", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE)
```

---

# Baseline (pre-intervention) trend & validation

The panels below use the **reference (`ref`)** arm as the pre-intervention
baseline, projected over the simulation horizon (**`r sim_horizon`**, with
`r base_year` the seeded jump-off year), and — where the GBD artifact is
available — **complemented with GBD 2023 historical series from 2000** so the
reader can judge continuity and validity at the simulation start. Following the
report convention, **solid lines are historical (GBD 2000–2023)** and **dashed
lines are model-projected (`r proj_years`)**; model *rates* are crude rates
(count ÷ total population) computed on the same basis as the GBD crude rates, so
the two are directly comparable at the join.

```{r baseline-aggregates, message=FALSE, warning=FALSE, echo=FALSE}
ref <- wsd[scenario == "ref"]

# Model annual aggregates (sum over age & sex; pop is identical across arms).
mod_pop <- ref[, .(pop = sum(pop)), by = year]
mod_yr  <- ref[, .(inc = sum(newcases), prev = sum(sick), rhd_deaths = sum(dead),
                   all_deaths = sum(all.mx)), by = year]
mod_yr  <- merge(mod_yr, mod_pop, by = "year")
mod_yr[, `:=`(inc_rate  = inc / pop * 1000,
              prev_rate = prev / pop * 1000,
              rhd_mort  = rhd_deaths / pop * 1e5,
              all_mort  = all_deaths / pop * 1e5)]

# GBD historical aggregates + crude rates (guarded).
build_hist <- function(cause, measure, per = 1000) {
  cnt <- gbd_count_yr(cause, measure)
  if (is.null(cnt) || is.null(gbd_pop_yr)) return(NULL)
  h <- merge(cnt, gbd_pop_yr, by = "year")
  h[, rate := count / pop * per]
  h[year >= 2000][order(year)]
}
h_inc   <- build_hist(CAUSE, "Incidence", 1000)
h_prev  <- build_hist(CAUSE, "Prevalence", 1000)
h_rhdmx <- build_hist(CAUSE, "Deaths", 1e5)
h_allmx <- build_hist("All causes", "Deaths", 1e5)

# Combine one historical + one projected series into a tidy DT for plotting.
combine_series <- function(hist_dt, hist_val, mod_dt, mod_val) {
  parts <- list()
  if (!is.null(hist_dt))
    parts[[1]] <- data.table(year = hist_dt$year, value = hist_dt[[hist_val]],
                             period = "Historical (GBD)")
  parts[[2]] <- data.table(year = mod_dt$year, value = mod_dt[[mod_val]],
                           period = "Projected (model)")
  out <- rbindlist(parts)
  out[, period := factor(period, levels = c("Historical (GBD)", "Projected (model)"))]
  out[]
}

lt_vals <- c("Historical (GBD)" = "solid", "Projected (model)" = "dashed")

plot_series <- function(dt, title, subtitle, ylab, color, ylab_comma = TRUE) {
  p <- ggplot(dt, aes(year, value, linetype = period)) +
    geom_line(linewidth = 1.1, color = color) +
    scale_linetype_manual(values = lt_vals, name = NULL) +
    labs(title = title, subtitle = subtitle, x = "Year", y = ylab) +
    base_theme() + theme(legend.position = "bottom")
  if (ylab_comma) p <- p + scale_y_continuous(labels = comma, limits = c(0, NA))
  p
}
```

## Incident cases and incidence rate

```{r baseline-incidence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.3}
d_inc_cnt  <- combine_series(h_inc, "count", mod_yr, "inc")
d_inc_rate <- combine_series(h_inc, "rate",  mod_yr, "inc_rate")

print(plot_series(d_inc_cnt,
  "Incident RHD cases per year — reference",
  paste0(loc1, "; solid = GBD historical, dashed = model-projected"),
  "New RHD cases per year", "#2166ac"))

print(plot_series(d_inc_rate,
  "RHD incidence rate — reference",
  "New RHD cases per 1,000 population per year",
  "Incident cases per 1,000", "#2166ac", ylab_comma = FALSE))
```

## Prevalent cases and prevalence

```{r baseline-prevalence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.3}
d_prev_cnt  <- combine_series(h_prev, "count", mod_yr, "prev")
d_prev_rate <- combine_series(h_prev, "rate",  mod_yr, "prev_rate")

print(plot_series(d_prev_cnt,
  "Prevalent RHD cases — reference",
  paste0(loc1, "; solid = GBD historical, dashed = model-projected"),
  "Prevalent RHD cases", "#b2182b"))

print(plot_series(d_prev_rate,
  "RHD prevalence — reference",
  "Prevalent RHD cases per 1,000 population",
  "Prevalent cases per 1,000", "#b2182b", ylab_comma = FALSE))
```

## Mortality rates — RHD and all causes

```{r baseline-mortality, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.3}
d_rhd_mort <- combine_series(h_rhdmx, "rate", mod_yr, "rhd_mort")
d_all_mort <- combine_series(h_allmx, "rate", mod_yr, "all_mort")

print(plot_series(d_rhd_mort,
  "RHD mortality rate — reference",
  "RHD cause-specific deaths per 100,000 population",
  "RHD deaths per 100,000", "#4d4d4d", ylab_comma = FALSE))

print(plot_series(d_all_mort,
  "All-cause mortality rate — reference",
  "All-cause deaths per 100,000 population (RHD + background)",
  "All-cause deaths per 100,000", "#333333", ylab_comma = FALSE))
```

```{r validation-note, echo=FALSE, results='asis'}
if (have_gbd && have_pop) {
  # Continuity check: model base-year prevalence rate vs last GBD prevalence rate.
  m0 <- mod_yr[year == base_year, prev_rate]
  g0 <- if (!is.null(h_prev)) h_prev[year == max(year), rate] else NA_real_
  if (length(m0) && is.finite(g0))
    cat(sprintf(
"The model is seeded at %d from GBD 2023 prevalence, so the projected prevalence
rate at the jump-off year (**%.2f per 1,000**) sits close to the last observed
GBD value (**%.2f per 1,000** in %d), giving visual continuity across the
2024 seam. The RHD incidence array carries a %.1f%%/year secular decline, so
incidence drifts down over the horizon while prevalence reflects the balance of
incidence, progression and improving survival.",
      base_year, m0, g0, h_prev[year == max(year), year],
      100 * (1 - meta$incidence_trend)))
} else {
  cat("GBD historical artifact unavailable — historical overlays were skipped and",
      "the panels above show the model-projected reference series only.")
}
```

## Age- and sex-specific profiles (every 10 years, 2000–2100)

The figures below show age profiles at **decade years from 2000 to 2100**, split
by sex. Years through 2023 are drawn from GBD (age-group rates); years from
`r base_year` on are model-projected, folded into the same GBD age groups so the
two share one comparable age axis.

```{r age-profiles, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7.5}
dec_years <- seq(2000, 2100, by = 10)

# Model age profile by (year, sex, GBD group): prevalence per 1,000, RHD mort /100k.
refg <- copy(ref)
refg[, grp := age_to_grp(age)]
mod_prof <- refg[, .(prev = sum(sick), dead = sum(dead), pop = sum(pop)),
                 by = .(year, sex, grp)]
mod_prof[, `:=`(prev_rate = prev / pop * 1000, mort = dead / pop * 1e5)]
mod_prof <- mod_prof[year %in% dec_years]

# GBD age profile from the Rate metric (per-age-group rate).
if (have_gbd) {
  g_prev <- gbd[cause_name == CAUSE & measure_name == "Prevalence" & metric_name == "Rate",
                .(year, sex = sex_name, grp = age_name, prev_rate = val / 100)]  # /100k*1000
  g_mort <- gbd[cause_name == CAUSE & measure_name == "Deaths" & metric_name == "Rate",
                .(year, sex = sex_name, grp = age_name, mort = val)]             # per 100k
  g_prev <- g_prev[year %in% dec_years]
  g_mort <- g_mort[year %in% dec_years]
} else {
  g_prev <- g_mort <- NULL
}

build_profile <- function(mod_col, gbd_dt, gbd_col) {
  parts <- list(mod_prof[, .(year, sex, grp, value = get(mod_col))][year >= base_year])
  if (!is.null(gbd_dt))
    parts <- c(list(gbd_dt[year < base_year, .(year, sex, grp, value = get(gbd_col))]), parts)
  out <- rbindlist(parts)
  out <- out[grp %in% names(grp_mid)]
  out[, age_mid := grp_mid[grp]]
  out[, sex := factor(sex, levels = names(sex_cols))]
  out[order(year, sex, age_mid)]
}

prof_prev <- build_profile("prev_rate", g_prev, "prev_rate")
prof_mort <- build_profile("mort",      g_mort, "mort")

ggplot(prof_prev, aes(age_mid, value, color = sex)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ year, ncol = 3, scales = "free_y") +
  scale_color_manual(values = sex_cols, name = NULL) +
  labs(title = "RHD prevalence by age and sex, every 10 years",
       subtitle = "Prevalent cases per 1,000; GBD (≤2023) and model-projected (≥ jump-off) share one age axis",
       x = "Age (years)", y = "Prevalent cases per 1,000") +
  base_theme(12) + theme(legend.position = "bottom")
```

```{r age-profiles-mort, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7.5}
ggplot(prof_mort, aes(age_mid, value, color = sex)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ year, ncol = 3, scales = "free_y") +
  scale_color_manual(values = sex_cols, name = NULL) +
  labs(title = "RHD mortality rate by age and sex, every 10 years",
       subtitle = "RHD deaths per 100,000; GBD (≤2023) and model-projected (≥ jump-off)",
       x = "Age (years)", y = "RHD deaths per 100,000") +
  base_theme(12) + theme(legend.position = "bottom")
```

---

# Results / outputs {#results}

The headline outputs contrast the two arms across three outcomes — **incident
RHD**, **total (prevalent) RHD** and **RHD deaths** — following the two-row
layout of the *Lancet Global Health* investment-case figure: **absolute counts
per year (both arms) on top**, and the **relative reduction in the rate (%)** on
the bottom. Reductions are computed in-report as **`ref − sap`**.

```{r results-compute, message=FALSE, warning=FALSE, echo=FALSE}
by_scn <- wsd[, .(inc = sum(newcases), prev = sum(sick), deaths = sum(dead),
                  pop = sum(pop)), by = .(scenario, year)]
by_scn[, `:=`(inc_rate = inc / pop * 1000, prev_rate = prev / pop * 1000,
              mort_rate = deaths / pop * 1e5)]

w2 <- dcast(by_scn, year ~ scenario,
            value.var = c("inc", "prev", "deaths", "inc_rate", "prev_rate", "mort_rate"))
# Absolute reductions (counts) and relative reduction in the rate (%).
w2[, `:=`(
  red_inc    = inc_ref - inc_sap,
  red_prev   = prev_ref - prev_sap,
  red_deaths = deaths_ref - deaths_sap,
  pct_inc    = fifelse(inc_rate_ref > 0,  (inc_rate_ref - inc_rate_sap)   / inc_rate_ref  * 100, 0),
  pct_prev   = fifelse(prev_rate_ref > 0, (prev_rate_ref - prev_rate_sap) / prev_rate_ref * 100, 0),
  pct_deaths = fifelse(mort_rate_ref > 0, (mort_rate_ref - mort_rate_sap) / mort_rate_ref * 100, 0))]

oc_levels <- c("Incident RHD", "Total (prevalent) RHD", "RHD deaths")
```

## Figure — two-row impact panel (reference vs SAP scale-up)

```{r fig-lancet, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7}
# Top row: absolute counts per year, both arms.
abs_long <- rbindlist(list(
  by_scn[, .(year, outcome = "Incident RHD",          value = inc,    series = as.character(scenario))],
  by_scn[, .(year, outcome = "Total (prevalent) RHD", value = prev,   series = as.character(scenario))],
  by_scn[, .(year, outcome = "RHD deaths",            value = deaths, series = as.character(scenario))]))
abs_long[, row := "Absolute count (n / year)"]

# Bottom row: relative reduction in the rate (%), single series.
pct_long <- rbindlist(list(
  w2[, .(year, outcome = "Incident RHD",          value = pct_inc)],
  w2[, .(year, outcome = "Total (prevalent) RHD", value = pct_prev)],
  w2[, .(year, outcome = "RHD deaths",            value = pct_deaths)]))
pct_long[, `:=`(series = "reduction", row = "Relative reduction in rate (%)")]

combo <- rbind(abs_long, pct_long)
combo[, outcome := factor(outcome, levels = oc_levels)]
combo[, series  := factor(series, levels = c("ref", "sap", "reduction"))]
# Panel ordering: row-major so the three "count" panels form the top row.
panel_lv <- c(paste0(oc_levels, "\nAbsolute count (n / year)"),
              paste0(oc_levels, "\nRelative reduction in rate (%)"))
combo[, panel := factor(paste0(outcome, "\n", row), levels = panel_lv)]

col_series <- c(ref = scenario_cols[["ref"]], sap = scenario_cols[["sap"]],
                reduction = reduction_col)
lab_series <- c(ref = "Reference", sap = "SAP scale-up", reduction = "Reduction (ref − sap)")

ggplot(combo[year >= base_year], aes(year, value, color = series)) +
  geom_hline(data = combo[grepl("%", panel), .(panel = unique(panel))], aes(yintercept = 0),
             linetype = "dashed", color = "grey60", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  facet_wrap(~ panel, ncol = 3, scales = "free_y") +
  scale_color_manual(values = col_series, labels = lab_series, name = NULL,
                     breaks = c("ref", "sap", "reduction")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Projected RHD impact: reference vs SAP scale-up",
       subtitle = paste0("Top: absolute counts per year (both arms). Bottom: ",
                         "relative reduction in the rate (%). ", loc1, ", ", proj_years),
       x = "Year", y = NULL) +
  base_theme(12) + theme(legend.position = "bottom")
```

```{r results-interpretation, echo=FALSE, results='asis'}
red_d_end  <- w2[year == year_max, red_deaths]
pct_d_end  <- w2[year == year_max, pct_deaths]
pct_p_end  <- w2[year == year_max, pct_prev]
cat(sprintf(
"By construction, **incidence is essentially identical across the two arms**
(`eff_ir = 1`): the incident-RHD panels overlap and the relative reduction is
~0%%. SAP bites on **mortality** — by %d the scale-up averts about **%s RHD
deaths in that single year** (a **%s** reduction in the RHD death rate). Because
those averted deaths mean more people remain alive with RHD (predominantly in
the earlier stages), modelled **prevalent cases are slightly *higher* under the
scale-up** (a %s change in the prevalence rate by %d), which appears as a small
*negative* reduction in the prevalent-RHD panel. No probabilistic uncertainty
bands are shown: the pipeline is deterministic and runs no probabilistic
sensitivity analysis, so none are available.",
  year_max, fmt_n(red_d_end), fmt_pct(pct_d_end/100, 1),
  fmt_pct(pct_p_end/100, 1), year_max))
```

## Annual and cumulative RHD deaths averted

```{r fig-deaths-averted, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.3}
avert <- w2[, .(year, annual = red_deaths)]
avert[, cumulative := cumsum(annual)]

print(ggplot(avert[year >= base_year], aes(year, annual)) +
  geom_area(fill = "#2166ac", alpha = 0.25) +
  geom_line(linewidth = 1.1, color = "#2166ac") +
  scale_y_continuous(labels = comma) +
  labs(title = "Annual RHD deaths averted (reference − SAP scale-up)",
       subtitle = paste0(loc1, "; benefit grows as SAP coverage ramps to target"),
       x = "Year", y = "RHD deaths averted per year") +
  base_theme())

print(ggplot(avert[year >= base_year], aes(year, cumulative)) +
  geom_line(linewidth = 1.2, color = "#b2182b") +
  scale_y_continuous(labels = comma) +
  labs(title = "Cumulative RHD deaths averted",
       subtitle = paste0("Running total vs reference, ", proj_years),
       x = "Year", y = "Cumulative RHD deaths averted") +
  base_theme())
```

## Summary tables — cumulative impact by scenario

```{r tbl-impact, message=FALSE, warning=FALSE, echo=FALSE}
# Cumulative flows over the horizon + prevalent stock at the final year.
cum_by <- wsd[, .(inc = sum(newcases), deaths = sum(dead)), by = scenario]
prev_end <- wsd[year == year_max, .(prev = sum(sick)), by = scenario]
cum_by <- merge(cum_by, prev_end, by = "scenario")

get_v <- function(dt, col, scn) dt[scenario == scn, get(col)]
impact_tab <- data.table(
  Outcome = c("Cumulative incident RHD cases",
              paste0("Prevalent RHD cases (", year_max, ")"),
              "Cumulative RHD deaths"),
  Reference = c(get_v(cum_by, "inc", "ref"), get_v(cum_by, "prev", "ref"), get_v(cum_by, "deaths", "ref")),
  `SAP scale-up` = c(get_v(cum_by, "inc", "sap"), get_v(cum_by, "prev", "sap"), get_v(cum_by, "deaths", "sap")))
impact_tab[, `Averted (ref − sap)` := Reference - `SAP scale-up`]

impact_disp <- copy(impact_tab)
num_cols <- c("Reference", "SAP scale-up", "Averted (ref − sap)")
impact_disp[, (num_cols) := lapply(.SD, function(x) fmt_n(x, 0)), .SDcols = num_cols]

kable(impact_disp,
      caption = paste0("Table 3. Cumulative RHD impact by scenario over ", proj_years,
                       " (", loc1, "). Incidence is unchanged by design; deaths are ",
                       "averted; prevalent cases rise as survival improves."),
      align = c("l", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(3, bold = TRUE, background = "#eef2f7") |>
  footnote(general = paste0(
    "Averted = reference − scale-up. A negative 'averted' for prevalent cases ",
    "means the scale-up leaves MORE people alive with RHD (a survival benefit), ",
    "not a harm."))
```

```{r tbl-impact-dt, message=FALSE, warning=FALSE, echo=FALSE}
# Interactive annual series: deaths by arm + deaths averted.
annual_dt <- dcast(by_scn, year ~ scenario, value.var = "deaths")
setnames(annual_dt, c("ref", "sap"), c("RHD deaths (ref)", "RHD deaths (sap)"), skip_absent = TRUE)
annual_dt[, `Deaths averted` := `RHD deaths (ref)` - `RHD deaths (sap)`]
annual_dt[, `Cumulative averted` := cumsum(`Deaths averted`)]
datatable(annual_dt, rownames = FALSE,
          caption = "Annual RHD deaths by arm and deaths averted (searchable / sortable).",
          options = list(pageLength = 12, order = list(list(0, "asc")))) |>
  formatRound(c("RHD deaths (ref)", "RHD deaths (sap)", "Deaths averted",
                "Cumulative averted"), digits = 0)
```

## RHD stage composition and the surgery service

Living RHD is resolved into the four WHF stages. The stacked areas below show how
the prevalent pool is distributed across **A** (minimal), **B** (mild), **C**
(advanced without complications) and **D** (advanced with complications) over the
horizon, in each arm; the compiled stage table `rhd_stage_model_long` provides
these per-stage prevalences directly.

```{r fig-stage-composition, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.6}
stage_ty <- stage_long[, .(cases = sum(cases)), by = .(scenario, year, stage)]
ggplot(stage_ty[year >= base_year], aes(year, cases, fill = stage)) +
  geom_area(alpha = 0.9, colour = "white", linewidth = 0.1) +
  facet_wrap(~ scenario, ncol = 2, labeller = as_labeller(scenario_labels_short)) +
  scale_fill_manual(values = stage_cols, labels = stage_labels, name = "WHF stage") +
  scale_y_continuous(labels = comma) +
  labs(title = "Prevalent RHD by WHF stage (A/B/C/D)",
       subtitle = paste0(loc1, ", ", proj_years,
                         "; the living RHD pool resolved into the four stages"),
       x = "Year", y = "Prevalent RHD cases") +
  base_theme(12) + theme(legend.position = "bottom")
```

Surgery is modelled as a **clinical service**: each year a fixed fraction of the
stage C and D stocks requiring surgery is operated on (at the prevailing surgery
coverage). The service volume enters the cost model and reduces C → D progression
and D → RHD death, but it never removes patients from a stage. The trace below
comes from the annual-flow table `rhd_annual_flows`.

```{r fig-surgery-trace, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4}
surg_ty <- flows[, .(surgeries_C = sum(surgeries_C), surgeries_D = sum(surgeries_D)),
                 by = .(scenario, year)]
surg_lg <- melt(surg_ty[scenario == "sap"], id.vars = c("scenario", "year"),
                measure.vars = c("surgeries_C", "surgeries_D"),
                variable.name = "stage", value.name = "n")
surg_lg[, stage := factor(fifelse(stage == "surgeries_C", "C", "D"), levels = c("C", "D"))]
ggplot(surg_lg[year >= base_year], aes(year, n, fill = stage)) +
  geom_area(alpha = 0.9, colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = c(C = stage_cols[["C"]], D = stage_cols[["D"]]),
                    labels = c(C = "Surgeries in stage C", D = "Surgeries in stage D"),
                    name = NULL) +
  scale_y_continuous(labels = comma) +
  labs(title = "Surgical volume delivered per year (a service, not a state)",
       subtitle = paste0("SAP arm; surgery coverage is held equal in both arms, so the ",
                         "reference volume is near-identical"),
       x = "Year", y = "Surgeries delivered per year") +
  base_theme() + theme(legend.position = "bottom")
```

---

# Economic evaluation {#economic-evaluation}

```{r econ-guard, echo=FALSE, results='asis'}
if (!have_econ)
  cat("> **Economic artifacts unavailable.** `output/tables/rhd_economic_results.rds`",
      "was not found, so this section is skipped. Run `scripts/08_economic_evaluation.R`",
      "to generate the economics.")
```

All economics are produced by `08_economic_evaluation.R` and read here from
`rhd_economic_results.rds` (`$summary`, `$budget_impact`, `$incremental`,
`$params`); the same values are mirrored in `rhd_economic_summary.csv` and
`rhd_budget_impact.csv`. Costs are applied to the model's stage-and-flow
volumes — **echo screening** on the total population screened, **optimal
treatment (SAP)** on the living-RHD stock on treatment, and **surgery** on the
annual surgery trace — with RHD deaths driving the monetised benefit;
**incidence is not costed**. Formally the annual programme cost is
$c_{\text{screen}} = \sum n_{\text{screened}}\times \text{cost}_{\text{screen}}$,
$c_{\text{sap}} = \sum n_{\text{on optimal treatment}}\times \text{cost}_{\text{sap}}$,
$c_{\text{surgery}} = \sum n_{\text{surgeries}}\times \text{cost}_{\text{surgery}}$.

## Method and discounting

```{r econ-method, echo=FALSE, results='asis'}
if (have_econ) {
  cat(sprintf(
"Costs and benefits are **discounted at %s per year** to a base year of **%d**
with discount factor $\\text{df}_t = (1+r)^{-(t-%d)}$, $r=%.2f$. Annual incremental
cost is $\\text{cost}(\\text{sap}) - \\text{cost}(\\text{ref})$. Annual RHD
**deaths averted** ($\\text{ref}-\\text{sap}$) are monetised as
$\\text{deaths averted}\\times(\\text{VSL}_t + \\text{GDPpc}_t)$, where
$\\text{GDPpc}_t = \\%s\\times(1+%.2f)^{(t-%d)}$ (grown from the base-year value)
and $\\text{VSL}_t = %g\\times\\text{GDPpc}_t$. The **benefit–cost ratio** is
$\\sum_t \\text{df}_t\\,\\text{benefit}_t \\big/ \\sum_t \\text{df}_t\\,\\text{cost}_t$;
**cost per DALY averted** is discounted cost divided by discounted deaths averted
times %g DALYs per death.",
    fmt_pct(eparams$disc_rate, 0), disc_base, disc_base, eparams$disc_rate,
    fmt_usd(eparams$gdp_pc_base), eparams$gdp_growth, eparams$gdp_pc_base_year,
    eparams$vsl_mult, eparams$dalys_per_death))
}
```

$$
\text{BCR} = \frac{\sum_t \text{df}_t\,\text{benefit}_t}{\sum_t \text{df}_t\,\text{cost}_t},
\qquad
\text{cost/DALY} = \frac{\sum_t \text{df}_t\,\text{inc\_cost}_t}{\big(\sum_t \text{df}_t\,\text{deaths\_averted}_t\big)\times \text{DALYs/death}} .
$$

## Economic summary

```{r tbl-econ-summary, message=FALSE, warning=FALSE, echo=FALSE}
if (have_econ) {
  es <- copy(econ_sum)
  es_disp <- data.table(
    Location = es$location,
    `Incremental cost` = vapply(es$total_cost, usd_m, character(1)),
    `Monetised benefit` = vapply(es$total_benefit, usd_m, character(1)),
    `Net benefit` = vapply(es$net_benefit, usd_m, character(1)),
    BCR = formatC(es$bcr, format = "f", digits = 2),
    `Deaths averted` = fmt_n(es$deaths_averted, 0),
    `Cost / death averted` = vapply(es$cost_per_death, fmt_usd, character(1)),
    `Cost / DALY averted` = vapply(es$cost_per_daly, fmt_usd, character(1)))
  kbl <- kable(es_disp,
        caption = paste0("Table 4. Economic summary (discounted, base-year US$). ",
                         "Incremental cost, monetised benefit, net benefit, BCR, ",
                         "and cost-effectiveness per location and TOTAL."),
        align = c("l", rep("r", 7))) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 12) |>
    row_spec(0, bold = TRUE)
  tot_row <- which(es$location == "TOTAL")
  if (length(tot_row)) kbl <- row_spec(kbl, tot_row, bold = TRUE, background = "#eef2f7")
  kbl <- footnote(kbl, general = paste0(
    "Benefit = deaths averted × (VSL + GDP-pc productivity), discounted at ",
    fmt_pct(eparams$disc_rate, 0), " to ", disc_base,
    ". Cost/DALY uses ", eparams$dalys_per_death, " DALYs per RHD death."))
  kbl
}
```

## Budget impact

```{r tbl-budget, message=FALSE, warning=FALSE, echo=FALSE}
if (have_econ) {
  # Compact kable at decade years; full interactive table below.
  show_years <- intersect(seq(base_year, year_max, by = 10), budget$year)
  bshow <- budget[year %in% show_years,
                  .(Year = year,
                    `Cost — reference` = cost_ref,
                    `Cost — SAP scale-up` = cost_sap,
                    `Incremental cost` = incremental,
                    `Deaths averted` = deaths_avert)]
  for (cc in c("Cost — reference", "Cost — SAP scale-up", "Incremental cost"))
    bshow[, (cc) := vapply(get(cc), usd_m, character(1))]
  bshow[, `Deaths averted` := fmt_n(`Deaths averted`, 0)]
  kable(bshow,
        caption = paste0("Table 5. Budget impact at decade years (undiscounted ",
                         "annual programme cost by arm, incremental cost, and ",
                         "RHD deaths averted)."),
        align = c("l", rep("r", 4))) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 12) |>
    row_spec(0, bold = TRUE)
}
```

```{r dt-budget, message=FALSE, warning=FALSE, echo=FALSE}
if (have_econ) {
  bdt <- budget[, .(Year = year,
                    `Cost ref` = round(cost_ref), `Cost sap` = round(cost_sap),
                    `Incremental` = round(incremental),
                    `Deaths averted` = round(deaths_avert, 1),
                    `Screen` = round(c_screen), `SAP` = round(c_sap),
                    `Surgery` = round(c_surgery))]
  datatable(bdt, rownames = FALSE,
            caption = paste0("Full annual budget-impact table (", proj_years,
                             "). Cost columns in base-year US$; Screen/SAP/Surgery ",
                             "are the SAP-arm component costs."),
            options = list(pageLength = 12, scrollX = TRUE,
                           order = list(list(0, "asc")))) |>
    formatCurrency(c("Cost ref", "Cost sap", "Incremental", "Screen", "SAP", "Surgery"),
                   currency = "$", digits = 0) |>
    formatRound("Deaths averted", digits = 1)
}
```

## Figure 3 — Cost of interventions scaled to target coverage, by source

```{r fig3-cost-by-source, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
if (have_econ) {
  cost_src <- melt(
    budget[, .(year, c_screen, c_sap, c_surgery)],
    id.vars = "year", variable.name = "component", value.name = "cost")
  comp_lab <- c(c_screen = "Echo screening", c_sap = "Secondary prophylaxis (SAP)",
                c_surgery = "Valve surgery")
  comp_col <- c(c_screen = "#4E79A7", c_sap = "#F28E2B", c_surgery = "#E15759")
  cost_src[, component := factor(component, levels = names(comp_lab))]

  ggplot(cost_src, aes(year, cost / 1e6, fill = component)) +
    geom_area(alpha = 0.9, colour = "white", linewidth = 0.1) +
    annotate("rect", xmin = ramp_start, xmax = ramp_end, ymin = -Inf, ymax = Inf,
             alpha = 0.06, fill = "black") +
    scale_fill_manual(values = comp_col, labels = comp_lab, name = "Cost component") +
    scale_y_continuous(labels = comma) +
    labs(title = "Figure 3. Cost of the SAP scale-up scaled to target coverage, by source",
         subtitle = paste0("SAP-arm annual programme cost (US$ millions) by component; ",
                           "shaded = scale-up ramp (", ramp_start, "–", ramp_end, ")"),
         x = "Year", y = "Annual cost (US$ millions)",
         caption = "Components are the SAP-arm totals from rhd_budget_impact.csv; sum to the SAP-arm total cost.") +
    base_theme() + theme(legend.position = "bottom")
}
```

## Benefit versus cost, and net benefit over time

```{r fig-benefit-cost, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
if (have_econ) {
  ic <- copy(incr)
  # Discounted annual streams.
  ic[, `:=`(disc_cost = inc_cost * df, disc_benefit = benefit_total * df)]
  bc_long <- rbindlist(list(
    ic[, .(year, value = disc_cost,    stream = "Discounted incremental cost")],
    ic[, .(year, value = disc_benefit, stream = "Discounted monetised benefit")]))

  print(ggplot(bc_long, aes(year, value / 1e6, color = stream)) +
    geom_line(linewidth = 1.1) +
    scale_color_manual(values = c("Discounted incremental cost" = "#E15759",
                                  "Discounted monetised benefit" = "#59A14F"),
                       name = NULL) +
    scale_y_continuous(labels = comma) +
    labs(title = "Annual discounted benefit vs cost",
         subtitle = paste0("Discounted to ", disc_base, " at ",
                           fmt_pct(eparams$disc_rate, 0), "; US$ millions"),
         x = "Year", y = "US$ millions (discounted)") +
    base_theme() + theme(legend.position = "bottom"))

  ic[, cum_net := cumsum((benefit_total - inc_cost) * df)]
  print(ggplot(ic, aes(year, cum_net / 1e9)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1.2, color = "#2166ac") +
    scale_y_continuous(labels = comma) +
    labs(title = "Cumulative discounted net benefit",
         subtitle = "Running total of (benefit − cost), discounted; US$ billions",
         x = "Year", y = "Cumulative net benefit (US$ billions)") +
    base_theme())
}
```

---

# Limitations and interpretation

- **Deterministic point estimates.** The pipeline runs a single deterministic
  scenario per arm with no probabilistic sensitivity analysis, so no uncertainty
  intervals are available; all clinical, efficacy, coverage and economic inputs
  (Table 1) are point values.
- **Deaths averted, not just delayed.** Because SAP acts on RHD-specific
  mortality in every living stage (case fatality), it genuinely removes RHD
  deaths from the horizon. A direct corollary is that **prevalent RHD rises
  slightly** under the scale-up (survivors live longer, predominantly in the
  earlier stages); this is a benefit, not a harm, and incidence is unchanged by
  design (`eff_ir = 1`).
- **Single location.** Only Indonesia is currently run; the engine and tables
  are location-general, but all results here are national.
- **Calibration and held-fixed structure.** Incidence and case fatality were
  calibrated (2000–2019) against GBD prevalence and all-cause deaths; background
  (non-RHD) mortality and the GBD age–sex rate pattern are **held fixed** at
  their base-year levels, with an explicit incidence secular trend
  (`r if (exists("meta")) sprintf("%.1f%%/year", 100*(1-meta$incidence_trend)) else "≈1.5%/year"`).
- **Stage split and stage parameters assumed.** GBD reports *total* RHD
  prevalence only; the A/B/C/D seed split (a multi-state severity study for
  A/B/C, normalised to the non-D share, plus the complicated-RHD fraction for D)
  is a tagged assumption. Within-RHD mortality is carried by the per-stage
  RHD-death probabilities rather than the calibrated case fatality (used only as
  a base-year anchor). The A/B/C/D transition, regression and per-stage mortality
  probabilities are `[CALIBRATE]` values and **stage-transition calibration
  against local echocardiographic prevalence targets is pending**, so absolute
  stage-specific levels are best read as structural illustrations rather than
  calibrated estimates.
- **SAP acts on stage-specific mortality, not incidence.** The modelled benefit
  operates entirely through reduced RHD-specific case fatality across the four
  stages; the intervention does not alter RHD incidence (`eff_ir = 1`), so it
  reduces deaths and shifts the prevalent stage mix rather than preventing new
  disease.
- **Surgery as a population-average service.** Surgery is represented as a fixed
  fraction of the stage C and D stocks operated on each year, reducing C → D
  progression and D → RHD-death risk in proportion to its coverage, rather than
  moving patients into a distinct post-operative stock; operative mortality is
  not modelled separately. Because surgery coverage and effects are held equal in
  both arms, surgery does not drive the incremental result.
- **Four-stage annual discretisation.** The continuous course of valve disease is
  discretised into four A/B/C/D stages advanced on a one-year cycle, with an open
  terminal 100+ age group that accumulates survivors; sub-annual dynamics,
  within-stage heterogeneity and very-old-age dynamics are approximate.
- **Economic assumptions.** The monetised benefit uses a VSL of
  `r if (have_econ) eparams$vsl_mult else 30`× GDP per capita plus a GDP-per-capita
  productivity term and `r if (have_econ) eparams$dalys_per_death else 30` DALYs
  per RHD death; unit costs are literature/calibrated values. The benefit–cost
  ratio is sensitive to the VSL multiplier and discount rate.
- **Historical validation is a crude-rate comparison.** GBD historical series
  use GBD counts over observed WPP population; the model rates are crude rates on
  the same basis. Neither is age-standardised, so shifts partly reflect the
  changing age structure.

---

# Reproducibility

```{r session-info, message=FALSE, warning=FALSE, echo=FALSE}
cat("Report generated:  ", format(Sys.time()), "\n")
cat("Project root:      ", wd, "\n")
cat("Model output files:", length(mfiles), "location file(s) from\n                    ",
    out_model_path, "\n")
cat("Location(s):       ", paste(LOCS, collapse = ", "), "\n")
cat("Scenarios:         ", paste(scenarios_present, collapse = ", "), "\n")
cat("Model horizon:     ", proj_years, "(seed / jump-off year:", base_year, ")\n")
cat("Artifacts loaded:\n")
cat("   baseline_state.rds        :", have_bs,   "\n")
cat("   rhd_stage_model_long.rds  :", have_stage_tab, "\n")
cat("   rhd_annual_flows.rds      :", have_flow_tab,  "\n")
cat("   rhd_economic_results.rds  :", have_econ, "\n")
cat("   GBD 2023 epi (historical) :", have_gbd,  if (have_gbd) paste0("  [", gbd_path, "]") else "", "\n")
cat("   pop_observed_1990_2024    :", have_pop,  "\n\n")
sessionInfo()
```
